Progress Report on One Simulated Dataset

### 2023-10-10 Docs/ProgressReportSim.Rmd
re-knit report
fix projection steps with FPCA parameters

- $t$ is equal-spaced on $[0, 1]$
- $f_0(t)=0$
- $\xi_k \sim N(0, \lambda_k)$, and $\lambda_k = 1, 0.5, 0.25, 0.125$ for k = 1, 2, 3, 4 respectively.

@@

# eigenvalues
lambda = 0.5^(0:(K-1))

# generated data 
## score
xi <- matrix(rnorm(N*K),N,K)
xi <- xi %*% diag(sqrt(lambda))

@@

# format into dataframe
# id = subject identifier (factor variable)
# sind = numeric value corresponding to the observed functional domain
# Y = functional response (binary)
# sind_inx = numeric value associated with order of "sind"
#            this is not necessary, but may be of convenience when you implement the method
df <- data.frame(id = factor(rep(1:N, each=J)),
                 t = rep(t, N), 
                 Y = as.vector(t(Y_i)),
                 eta_i = as.vector(t(eta_i)),
                 sind = rep(1:J, N))
# visualization
rand_id <- sample(N, size = 4)
df %>% filter(id %in% rand_id) %>% 
  ggplot()+
  geom_point(aes(x=sind, y=Y), size = 0.5)+
  geom_line(aes(x=sind, y=plogis(eta_i)), col = "red")+
  #geom_line(aes(x=sind_inx, y=eta_i), col = "blue")+
  facet_wrap(~id)
```

@@

@@

df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
df$bin <- as.numeric(as.character(df$bin))
# unique(df$bin)

df %>% 
  filter(id %in% rand_id) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  ggplot()+
  geom_point(aes(x=bin, y=num), size = 0.5)+
  facet_wrap(~id)+
  labs(x="Time", y = "Activity", title = "Number of active nimutes within each bin")
```

## Data split

- 400 subjects for training, 100 for out-of-sample prediction

```{r}
train_id <- sample(unique(df$id), size = N*0.8)
test_id <- setdiff(unique(df$id), train_id)

train_df <- df %>% filter(id %in% train_id)
test_df <- df %>% filter(id %in% test_id)
```

@@

- Used glmer with PIRLS (nAGQ=0) for local GLMMs step
- Fit on the training set

```{r, cache=TRUE}
# fit model on the training set
train_bin_lst <- split(train_df, f = train_df$bin)

# local GLMM and estimate latent function
# use PIRLS (nAGQ=0) to avoid near-unidentifiability issues 
t1=Sys.time()
df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
t2= Sys.time()
t_local_glmm <- t2-t1 
```

@@

# example estimated latent function
rand_id <- sample(train_id, 4)

df_est_latent %>% 
  filter(id %in% rand_id) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  mutate(eta_i = exp(eta_i)/(1+exp(eta_i))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, group = id, col = "estimated"))+
  geom_line(aes(x=sind, y=eta_i, group = id, col = "true"))+
  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

@@

```{r}
uni_eta_hat <- df_est_latent %>% filter(bin==sind)

@@

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
t2 <- Sys.time()
t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model
```

```{r}
plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean")
```

@@

plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
```

@@

- Use fastGFPCA::reeval_efunctions. Cannot install the package so used the source code directly
- Not sure how to decide the knots. I am using the bin boundaries for now (0, 10, 20, ...,1000), which is 101 knots.

@@

# number of knots
# not sure how to determine
# maybe for now use the bin endpoints? 
# knot_pos <- seq(0, J, by = 50)
knot_pos <- brks

# evaluate B-splines on binned grid
B <- spline.des(knots = knot_pos, x = mid, ord = p + 1,
                            outer.ok = TRUE)$design
# evaluate B-splines on original grid
Bnew <- spline.des(knots = knot_pos, x = 1:J, ord = p + 1,
                  outer.ok = TRUE)$design

# project binned eigenfunctions onto the original grid
efunctions_new <- matrix(NA, J, K)
for(k in 1:K){
    lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
    efunctions_new[,k] <- Bnew %*% coef(lm_mod)
}
```

@@

plot(mid, fpca_mod$efunctions[, 1], xlab="bin", ylab="PC1", pch=20, cex = 0.5)
lines(1:J, efunctions_new[, 1], col="blue")

plot(mid, fpca_mod$efunctions[, 2], xlab="bin", ylab="PC2", pch=20, cex = 0.5)
lines(1:J, efunctions_new[, 2], col="blue")

plot(mid, fpca_mod$efunctions[, 3], xlab="bin", ylab="PC3", pch=20, cex = 0.5)
lines(1:J, efunctions_new[, 3], col="blue")

plot(mid, fpca_mod$efunctions[, 4], xlab="bin", ylab="PC4", pch=20, cex = 0.5)
lines(1:J, efunctions_new[, 4], col="blue")

@@

- The projected eigenfunctions seem to be more wiggly than we would like?
- What is the reason for choosing projection over linear/spline interpolation?

#### Number of knots

- Try fewer knots?

```{r}
knot_pos <- seq(0, J, by = 100)

# evaluate B-splines on binned grid
B <- spline.des(knots = knot_pos, x = mid, ord = p + 1,
                            outer.ok = TRUE)$design
# evaluate B-splines on original grid
Bnew <- spline.des(knots = knot_pos, x = 1:J, ord = p + 1,
                  outer.ok = TRUE)$design

# project binned eigenfunctions onto the original grid
efunctions_new_11knots <- matrix(NA, J, K)
for(k in 1:K){
    lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
    efunctions_new_11knots[,k] <- Bnew %*% coef(lm_mod)
}
```

plot(mid, fpca_mod$efunctions[, 1], xlab="bin", ylab="PC1", pch=20, cex = 0.5)
lines(1:J, efunctions_new_11knots[, 1], col="blue")

plot(mid, fpca_mod$efunctions[, 2], xlab="bin", ylab="PC2", pch=20, cex = 0.5)
lines(1:J, efunctions_new_11knots[, 2], col="blue")

plot(mid, fpca_mod$efunctions[, 3], xlab="bin", ylab="PC3", pch=20, cex = 0.5)
lines(1:J, efunctions_new_11knots[, 3], col="blue")

plot(mid, fpca_mod$efunctions[, 4], xlab="bin", ylab="PC4", pch=20, cex = 0.5)
lines(1:J, efunctions_new_11knots[, 4], col="blue")

@@

#### Order of spline

- Try lower order? (order = 3)

```{r}
knot_pos <- brks

# evaluate B-splines on binned grid
B <- spline.des(knots = knot_pos, x = mid, ord = 3,
                            outer.ok = TRUE)$design
# evaluate B-splines on original grid
Bnew <- spline.des(knots = knot_pos, x = 1:J, ord = 3,
                  outer.ok = TRUE)$design

# project binned eigenfunctions onto the original grid
efunctions_new_2p <- matrix(NA, J, K)
for(k in 1:K){
    lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
    efunctions_new_2p[,k] <- Bnew %*% coef(lm_mod)
}
```

plot(mid, fpca_mod$efunctions[, 1], xlab="bin", ylab="PC1", pch=20, cex = 0.5)
lines(1:J, efunctions_new_2p[, 1], col="blue")

plot(mid, fpca_mod$efunctions[, 2], xlab="bin", ylab="PC2", pch=20, cex = 0.5)
lines(1:J, efunctions_new_2p[, 2], col="blue")

plot(mid, fpca_mod$efunctions[, 3], xlab="bin", ylab="PC3", pch=20, cex = 0.5)
lines(1:J, efunctions_new_2p[, 3], col="blue")

plot(mid, fpca_mod$efunctions[, 4], xlab="bin", ylab="PC4", pch=20, cex = 0.5)
lines(1:J, efunctions_new_2p[, 4], col="blue")

@@

### Debias with GLMM model

- The bias was caused by the misspecification of latent process (assume constant effect over smooth function). 
- What needs re-evaluation are eigenvalues (variance estimates). They will be used in Laplace Approximation for out-of-sample prediction. 
- We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction
- I will try to fit the model with projected eigenfunctions as above (with 101 knots). I am using mgcv::bam and setting method = "fREML" and discrete = TRUE to speed up computation

```{r}
df_phi <- data.frame(sind = 1:J, efunctions_new)
colnames(df_phi) <- c("sind", paste0("phi", 1:4))

# train_df$bin <- as.numeric(as.character(train_df$bin))
train_df <- train_df %>% left_join(df_phi, by = "sind")
train_df$id <- as.factor(train_df$id)
```

- Here, what if I replace "bin" with "t"? Should I regress of Y on t or sind?

```{r, class.source = "fold-show"}
# usethis::edit_r_environ()
t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(t, bs="cr", k=10)+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
t_debias <- t2-t1
```

- And now I try to extract the variance of scores (debiased eigenvalues). 
- Use the conclusion that connects smoothing parameters to variance

```{r, class.source='fold-show'}
# eigenvalues from FPCA
fpca_mod$evalues[1:4]
# transformed smoothing parameters
new_lambda <- 1/debias_glmm$sp[2:5]
new_lambda
# true eigenvalues
lambda
```

- Looks like the re-evaluated variance is still inflated from the true eigenvalues, but indeed doubled the result of FPCA.

>>>>

- $t$ is equal-spaced on $[0, 1]$
- $f_0(t)=0$
- $\xi_k \sim N(0, \lambda_k)$, and $\lambda_k = 1, 0.5, 0.25, 0.125$ for k = 1, 2, 3, 4 respectively. 
- In fact, since there is a constant factor $\sqrt{2}$ before each eigenfunction so that they are orthogonal

@@

# eigenvalues
lambda = 0.5^(0:(K-1)) 
```

```{r}
# generated training data 
## score
xi <- matrix(rnorm(N*K),N,K)
xi <- xi %*% diag(sqrt(lambda))

@@

# format into dataframe
# id = subject identifier (factor variable)
# sind = numeric value corresponding to the observed functional domain
# Y = functional response (binary)
# sind_inx = numeric value associated with order of "sind"
#            this is not necessary, but may be of convenience when you implement the method
train_df <- data.frame(id = factor(rep(1:N, each=J)),
                 t = rep(t, N), 
                 Y = as.vector(t(Y_i)),
                 eta_i = as.vector(t(eta_i)),
                 sind = rep(1:J, N))

# visualization
rand_id <- sample(N, size = 4)
train_df %>% filter(id %in% rand_id) %>% 
  ggplot()+
  geom_point(aes(x=sind, y=Y), size = 0.5)+
  geom_line(aes(x=sind, y=plogis(eta_i)), col = "red")+
  #geom_line(aes(x=sind_inx, y=eta_i), col = "blue")+
  facet_wrap(~id)
```

@@

- Training sample: N=500

@@

train_df$bin <- cut(train_df$sind, breaks = brks, include.lowest = T, labels = mid)
train_df$bin <- as.numeric(as.character(train_df$bin))
# unique(df$bin)

train_df %>% 
  filter(id %in% rand_id) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  ggplot()+
  geom_point(aes(x=bin, y=num), size = 0.5)+
  facet_wrap(~id)+
  labs(x="Time", y = "Activity", title = "Number of active nimutes within each bin")
```

@@

- Used glmer with PIRLS (nAGQ=0) for local GLMMs step
- Fit on the training set (N=500)
- Use original time index ([0, 1]) as independent variable

```{r, cache=TRUE}
# fit model on the training set
train_bin_lst <- split(train_df, f = train_df$bin)
# length(train_bin_lst)
# lapply(train_bin_lst, dim)

# local GLMM and estimate latent function
# use PIRLS (nAGQ=0) to avoid near-unidentifiability issues 
t1=Sys.time()
df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
t2= Sys.time()
t_local_glmm <- t2-t1 # less than 5 seconds
```

@@

# example estimated latent function
train_id <- unique(train_df$id)
rand_id <- sample(train_id, 4)

df_est_latent %>% 
  filter(id %in% rand_id) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  mutate(eta_i = exp(eta_i)/(1+exp(eta_i))) %>%
  ggplot()+
  geom_line(aes(x=t, y=eta_hat, group = id, col = "estimated"))+
  geom_line(aes(x=t, y=eta_i, group = id, col = "true"))+
  geom_point(aes(x=t, y = Y, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

@@

```{r}
uni_eta_hat <- df_est_latent %>% filter(bin==sind)
mid_t <- df_est_latent$t[df_est_latent$bin==df_est_latent$sind]
mid_t <- unique(mid_t) # the time points correspoinding to bin midpoints

@@

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid_t, var=T)
t2 <- Sys.time()
t_fpca <- t2-t1 # less than one second
```

- Estimated mean

```{r}
plot(mid_t, fpca_mod$mu, type = "l", xlab = "bin", ylab = "Mean")
```

- Estimated eigenfunctions

@@

plot(mid_t, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
plot(mid_t, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
plot(mid_t, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
plot(mid_t, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
```

- Correlation matrix

@@

<!-- https://bookdown.org/ssjackson300/Machine-Learning-Lecture-Notes/splines.html -->

- Use fastGFPCA::reeval_efunctions. Cannot install the package so used the source code directly
- I would like to adjust use time points on the [0, 1] scale as evaluation points
- Also, take number of knots and order from the FPCA model in step 3 (knots = 35, p = 3)
- Do the same kind of projection to mean functions as well

@@

# knots
knots <- 35
knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
knots_values <- knots_values * (max(mid_t) - min(mid_t)) + min(mid_t)

# evaluate B-splines on binned grid
B <- spline.des(knots = knots_values, x = mid_t, ord = p + 1,
                            outer.ok = TRUE)$design
# evaluate B-splines on original grid
Bnew <- spline.des(knots = knots_values, x = unique(train_df$t), ord = p + 1,
                  outer.ok = TRUE)$design

# project binned eigenfunctions onto the original grid
efunctions_new <- matrix(NA, J, K)
for(k in 1:K){
    lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
    efunctions_new[,k] <- Bnew %*% coef(lm_mod)
}

# project mean functions 
lm_mod <- lm(fpca_mod$mu ~ B-1)
mu_new <- Bnew %*% coef(lm_mod)
```

@@

plot(mid_t, fpca_mod$efunctions[, 1], xlab="bin", ylab="PC1", pch=20, cex = 0.5)
lines(t, efunctions_new[, 1], col="blue")

plot(mid_t, fpca_mod$efunctions[, 2], xlab="bin", ylab="PC2", pch=20, cex = 0.5)
lines(t, efunctions_new[, 2], col="blue")

plot(mid_t, fpca_mod$efunctions[, 3], xlab="bin", ylab="PC3", pch=20, cex = 0.5)
lines(t, efunctions_new[, 3], col="blue")

plot(mid_t, fpca_mod$efunctions[, 4], xlab="bin", ylab="PC4", pch=20, cex = 0.5)
lines(t, efunctions_new[, 4], col="blue")
```

```{r}
plot(mid_t, fpca_mod$mu, xlab="bin", ylab="Mean", pch=20, cex = 0.5)
lines(t, mu_new, col="blue")
```

- The projection seems to work very well!

### Debias with eigenvalues

- The bias was caused by the misspecification of latent process (assume constant effect over smooth function). 
- What needs re-evaluation are eigenvalues (variance estimates). They will be used in Laplace Approximation for out-of-sample prediction. 
- We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction. And out-of-sample scores are not estimated by FPCA but Laplace approximation.
- I will use the projected eigenfunctions from above. 
- I am using mgcv::bam and setting method = "fREML" and discrete = TRUE to speed up computation

```{r}
# dim(fpca_mod$efunctions)
df_phi <- data.frame(t = t, efunctions_new)
colnames(df_phi) <- c("t", paste0("phi", 1:4))

# train_df$bin <- as.numeric(as.character(train_df$bin))
train_df <- train_df %>% left_join(df_phi, by = "t")
train_df$id <- as.factor(train_df$id)
```

```{r, class.source = "fold-show"}
# usethis::edit_r_environ()
t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(t, bs="cr", k=10)+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
t_debias <- t2-t1 # less than 45 seconds
```

And now I try to extract the variance of scores (debiased eigenvalues), using the conclusion that connects smoothing parameters to variance

```{r, class.source='fold-show'}
# transformed smoothing parameters
new_lambda <- 1/debias_glmm$sp[2:5]
new_lambda
# new_lambda/lambda
# new_lambda/fpca_mod$evalues[1:4]
```

This eigenvalues, as well as the eigenfunctions, are actually not on the same scale of true values. By looking into the fpca.face source code, eigenfunctions are scaled by the square root of the number of total observations points. Since our model was fitted on the binnd grid, the actual eigenvaleus and eigenfunctions would be:

\begin{aligned}
\phi_k(t) & = \sqrt{n_{bin}}\hat{\phi}_k(t)  \\
\lambda_k & = \hat{\lambda}_k/n_{bin}

\end{aligned}

Let's rescale these quatities as below

- Re-scaled eigenvalues
```{r}
new_lambda/n_bin
```

- Re-scaled eigenfunctions

plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")

@@

- Both much closer to their true values

# Out-of-sample prediction with Laplace approximation

I will now estimate individual score of test sample using Laplace Approximation, and calculated the predicted track.

- Generate additional 200 subjects for out-of-sample performance evaluation
- Maximum observations time: 0.2, 0.4, 0.6, 0.8
- Prediction window: 0.2-0.4, 0.4-0.6, 0.6-0.8, 0.8-1.0

## Generate test data

```{r}
# generated testing data 
N_test <- 200
b_test <- matrix(rnorm(N_test*K), N_test, K) %*% diag(sqrt(lambda)) %*% t(phi) # of size N by J
# dim(b_test)

# latent gaussian function
eta_test <- t(vapply(1:N_test, function(x){
    f_0(x) + b_i[x,]
  }, numeric(J)))

# outcome binary function
Y_test <- matrix(rbinom(N_test*J, size=1, prob=plogis(eta_test)), 
                N_test, J, byrow=FALSE)

# format into dataframe
test_df <- data.frame(id = factor(rep(1:N_test, each=J)),
                 t = rep(t, N_test), 
                 Y = as.vector(t(Y_test)),
                 eta_i = as.vector(t(eta_test)),
                 sind = rep(1:J, N_test))
# bin
test_df$bin <- cut(test_df$sind, breaks = brks, include.lowest = T, labels = mid)
test_df$bin <- as.numeric(as.character(test_df$bin))
test_df$id <- as.factor(test_df$id)
test_id <- unique(test_df$id)
```

## Laplace approximation

```{r}
# the model
## now, with expanded mean and eigenfuncitons, a Bernoulli distribution at each time point
Model <- function(parm, Data){
  xi <- parm[Data$pos.xi]

  # log-prior
  xi.prior <- dmvnorm(xi, mean = rep(0, Data$K), sigma=Data$tao, log = TRUE)

  # log-posterior likelihood
  eta <- Data$f0+Data$X %*% xi
  p <- exp(eta)/(1+exp(eta))
  LL <- sum(dbern(x=Data$y, prob=p, log = TRUE)) # log likelihood of Y|xi
  LP <- LL+sum(xi.prior) # joint log likelihood of (Y, xi)

  # output
  Modelout <- list(LP=LP, Dev=-2*LL, Monitor=LL,
                   yhat=Data$y, parm=parm)
  return(Modelout)
}
```

```{r}
# parameter names
name_lst <- as.list(rep(0, K))
names(name_lst) <- paste("xi", 1:K, sep = "")
parm.names <- as.parm.names(name_lst)
pos.xi <- grep("xi", parm.names)
PGF <- function(Data) {
  xi <- rnorm(Data$K)
  return(xi)
}
mon.names <- "LP"
```

```{r, results='hide'}
# approximation process
# container
pred_list <- list()
# score <- array(NA, dim=c(length(test_id), K, 4))
# dimensions: subject, eigenfunction, maximum observation time

# prediction for a single subject
t1 <- Sys.time()
# per subject
for(i in seq_along(test_id)){
  df_i <- test_df %>% filter(id==test_id[i])

  # per max obs time
  for(tmax in c(0.2, 0.4, 0.6, 0.8)){
    df_it <- df_i %>% filter(t <= tmax)
    max_rid <- nrow(df_it)

    # into a list
    MyData <- list(K=K, PGF=PGF, 
                   X=efunctions_new[1: max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(new_lambda), f0=mu_new[1:max_rid, ])

    # fit laplace approximation
    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
    # Fit <- LaplacesDemon(Model, Data=MyData, Initial.Values = parm = rep(0, K),  Method = "BFGS")
    score <- Fit$Summary1[, "Mode"]

    # prediction
    eta_pred_out <- mu_new+efunctions_new%*%score
    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
  }

  # df_i$pred0.2[df_i$t<=0.2] <- NA
  # df_i$pred0.4[df_i$t<=0.4] <- NA
  # df_i$pred0.6[df_i$t<=0.6] <- NA
  # df_i$pred0.8[df_i$t<=0.8] <- NA

  pred_list[[i]] <- df_i
}
t2 <- Sys.time()
t_pred <- t2-t1 # About 3.5 minutes
```

```{r}
df_pred <- bind_rows(pred_list)

df_pred$pred0.2[df_pred$t<=0.2] <- NA
df_pred$pred0.4[df_pred$t<=0.4] <- NA
df_pred$pred0.6[df_pred$t<=0.6] <- NA
df_pred$pred0.8[df_pred$t<=0.8] <- NA

rand_test_id <- sample(test_id, 4)

p1<-df_pred %>%
  filter(id %in% rand_test_id) %>%
  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
  ggplot()+
  geom_point(aes(x=t, y=Y), size = 0.2)+
  geom_line(aes(x=t, y=eta_i, col = "True"))+
  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
  facet_wrap(~id)+
  labs(title = "Unscaled, debiased eigenvalues")

@@

### Compared to biased eigenvalues

```{r, results='hide'}
## With eigenvalues from FPCA
# container
pred_list2 <- list()

# per subject
for(i in seq_along(test_id)){
  df_i <- test_df %>% filter(id==test_id[i])

  # per max obs time
  for(tmax in c(0.2, 0.4, 0.6, 0.8)){
    df_it <- df_i %>% filter(t <= tmax)
    max_rid <- nrow(df_it)

    # into a list
    MyData <- list(K=K, PGF=PGF, 
                   X=efunctions_new[1: max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(fpca_mod$evalues[1:K]), f0=mu_new[1:max_rid, ])

    # fit laplace approximation
    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
    # Fit <- LaplacesDemon(Model, Data=MyData, Initial.Values = parm = rep(0, K),  Method = "BFGS")
    score <- Fit$Summary1[, "Mode"]

    # prediction
    eta_pred_out <- mu_new+efunctions_new%*%score
    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
  }

  # df_i$pred0.2[df_i$t<=0.2] <- NA
  # df_i$pred0.4[df_i$t<=0.4] <- NA
  # df_i$pred0.6[df_i$t<=0.6] <- NA
  # df_i$pred0.8[df_i$t<=0.8] <- NA

  pred_list2[[i]] <- df_i
}
```

```{r}
df_pred2 <- bind_rows(pred_list2)

df_pred2$pred0.2[df_pred2$t<=0.2] <- NA
df_pred2$pred0.4[df_pred2$t<=0.4] <- NA
df_pred2$pred0.6[df_pred2$t<=0.6] <- NA
df_pred2$pred0.8[df_pred2$t<=0.8] <- NA
# overview of predicted track
p2<-df_pred2 %>%
  filter(id %in% rand_test_id) %>%
  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
  ggplot()+
  geom_point(aes(x=t, y=Y), size = 0.2)+
  geom_line(aes(x=t, y=eta_i, col = "True"))+
  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
  facet_wrap(~id)+
  labs(title = "Unscaled, biased eigenvalues")

@@

### Scaled eigenfunctions and eigenvalues

- No need to scale mean functions

```{r, results='hide'}
## With debiased eigenvalues, also scale eigenvaleus and eigenfuncitons
# container
pred_list3 <- list()
efunctions_scaled <- efunctions_new*sqrt(n_bin)
evalues_scaled <- new_lambda/n_bin

# per subject
for(i in seq_along(test_id)){
  df_i <- test_df %>% filter(id==test_id[i])

  # per max obs time
  for(tmax in c(0.2, 0.4, 0.6, 0.8)){
    df_it <- df_i %>% filter(t <= tmax)
    max_rid <- nrow(df_it)

    # into a list
    MyData <- list(K=K, PGF=PGF, 
                   X=efunctions_scaled[1: max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(evalues_scaled), f0=mu_new[1:max_rid, ])

    # fit laplace approximation
    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
    # Fit <- LaplacesDemon(Model, Data=MyData, Initial.Values = parm = rep(0, K),  Method = "BFGS")
    score <- Fit$Summary1[, "Mode"]

    # prediction
    eta_pred_out <- mu_new+efunctions_scaled%*%score
    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
  }

  # df_i$pred0.2[df_i$t<=0.2] <- NA
  # df_i$pred0.4[df_i$t<=0.4] <- NA
  # df_i$pred0.6[df_i$t<=0.6] <- NA
  # df_i$pred0.8[df_i$t<=0.8] <- NA

  pred_list3[[i]] <- df_i
}
```

```{r}
df_pred3 <- bind_rows(pred_list3)

df_pred3$pred0.2[df_pred3$t<=0.2] <- NA
df_pred3$pred0.4[df_pred3$t<=0.4] <- NA
df_pred3$pred0.6[df_pred3$t<=0.6] <- NA
df_pred3$pred0.8[df_pred3$t<=0.8] <- NA
# overview of predicted track
p3 <- df_pred3 %>%
  filter(id %in% rand_test_id) %>%
  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
  ggplot()+
  geom_point(aes(x=t, y=Y), size = 0.2)+
  geom_line(aes(x=t, y=eta_i, col = "True"))+
  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
  facet_wrap(~id)+
  labs(title = "Scaled, unbiased eigenvalues")

```{r}
t_1sim <- (t_local_glmm+t_fpca+t_debias+t_pred)
```

- Time spent on one iteration: `r round(t_1sim, 2)` seconds
- Did debiased eigenvalues improve performance? The lines above look like so. I also wanna see the squared errors and AUC.
- Scaling eigenfunctions and values will change a prediction a little bit

```{r, fig.height=5, fig.width=15}
ggarrange(p1, p2, p3, nrow = 1, common.legend = T)
```

## ISE

```{r}
# calcualte ISE
err1 <- df_pred %>%
  mutate(err1 = (pred0.2-eta_i)^2,
         err2 = (pred0.4-eta_i)^2,
         err3 = (pred0.6-eta_i)^2,
         err4 = (pred0.8-eta_i)^2) %>%
  select(id, t, starts_with("err"))

err1$window = cut(err1$t, breaks = seq(0, 1, by = 0.2), include.lowest = T)
# table(err1$window)
err1 <- split(err1, f = err1$window)

tb1 <- lapply(err1, function(x){
  x %>%  group_by(id) %>% 
    summarise_at(vars(err1, err2, err3, err4), sum) %>%
    summarise_at(vars(err1, err2, err3, err4), mean)
}) %>% bind_rows(.id = "Window")
```

```{r}
# calcualte ISE
err2 <- df_pred2 %>%
  mutate(err1 = (pred0.2-eta_i)^2,
         err2 = (pred0.4-eta_i)^2,
         err3 = (pred0.6-eta_i)^2,
         err4 = (pred0.8-eta_i)^2) %>%
  select(id, t, starts_with("err"))

err2$window = cut(err2$t, breaks = seq(0, 1, by = 0.2), include.lowest = T)
# table(err1$window)
err2 <- split(err2, f = err2$window)

tb2 <- lapply(err2, function(x){
  x %>%  group_by(id) %>% 
    summarise_at(vars(err1, err2, err3, err4), sum) %>%
    summarise_at(vars(err1, err2, err3, err4), mean)
}) %>% bind_rows(.id = "Window")
```

```{r}
# calcualte ISE
err3 <- df_pred3 %>%
  mutate(err1 = (pred0.2-eta_i)^2,
         err2 = (pred0.4-eta_i)^2,
         err3 = (pred0.6-eta_i)^2,
         err4 = (pred0.8-eta_i)^2) %>%
  select(id, t, starts_with("err"))

err3$window = cut(err3$t, breaks = seq(0, 1, by = 0.2), include.lowest = T)
# table(err1$window)
err3 <- split(err3, f = err3$window)

tb3 <- lapply(err3, function(x){
  x %>%  group_by(id) %>% 
    summarise_at(vars(err1, err2, err3, err4), sum) %>%
    summarise_at(vars(err1, err2, err3, err4), mean)
}) %>% bind_rows(.id = "Window")
```

```{r}
# calcualte ISE
options(knitr.kable.NA = "")

ise <- full_join(tb1, tb2, by = "Window") %>% full_join(tb3, by = "Window")
colnames(ise) <-c("Window",  rep(seq(0.2, 0.8, by =0.2), 3))
ise %>%
  kable(digits = 3, table.attr = "style = \"color: black;\"") %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c(" "= 1, "Unscaled, debiased" = 4, "Unscaled, biased" = 4, "Scaled, debiased" = 4)) %>%
  add_header_above(c(" "=1, "Maximum observation time" = 12))
```

- Looks like the debiased eigenvalues helped.
- Scaling, on the other hand, does seem to help in general. 
- But shouldn't scaling improve performance? 
- Also, it seems like the longer the observed track, the closer the three cases are.

## AUC

```{r}
# calcualte AUC
auc_df1 <- df_pred %>%
  mutate(window = cut(t, breaks = seq(0, 1, by = 0.2), include.lowest = T)) %>%
  select(Y, starts_with("pred"), window)

obs_track <- seq(0.2, 0.8, by = 0.2)
pred_window <- unique(auc_df1$window)
auc_mat <- matrix(NA, length(pred_window), length(obs_track))
